---
title: "Starting with data"
author: "DC contributors and Dmytro Fishman"
minutes: 20
output: pdf_document
layout: topic
---

```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
```

------------

> ## Learning Objectives
>
> * load external data (CSV files) in memory using the temperature table
>  (`temperature.csv`) as an example
> * explore the structure and the content of a data frame in R
> * understand what factors are and how to manipulate them

------------

## Presentation of the Global Temperature Data

```{r, echo=FALSE, purl=TRUE}
### Presentation of the temperature data
```

We are studying the temperature. The dataset is stored as a CSV file: each row holds information for a
single measurement starting from 18 centuary and until now, and the columns represent:

| Column           | Description                        |
|------------------|------------------------------------|
| record\_id       | Unique id for the measurement      |
| month            | month of measurement               |
| day              | day of measurement                 |
| year             | year of measurement                |
| AverageTemperatureFahr         | regional average land temperature             |
| AverageTemperatureUncertaintyFahr      | the 95% confidence interval around the average                      |
| City              | city where temperature was measured           |
| country\_id | three first capitalised letters in the name of the country       |
| Country           | country where temperature was measured      |
| Latitude            | angular distance, of a point north or south from equator|
| Longitude          | angular distance, of a point east or west from Greenwich|

We are going to use the R function `download.file()` to download the CSV file
that contains the survey data from figshare, and we will use `read.csv()` to
load into memory (as a `data.frame`) the content of the CSV file.

To download the data into the `data/` subdirectory, do:

```{r, eval=FALSE, purl=TRUE}
download.file("https://s3-eu-west-1.amazonaws.com/pfigshare-u-files/4938964/temperature.csv",
              "data/temperature.csv")
```

You are now ready to load the data:

```{r, eval=TRUE,  purl=FALSE}
temperature <- read.csv('data/temperature.csv')
```

This statement doesn't produce any output because, as you might recall,
assignment doesn't display anything. If we want to check that our data has been
loaded, we can print the variable's value: `temperature`. Alternatively, wrapping an
assignment in parentheses will perform the assignment and display it at the same
time.

```{r, eval = TRUE, purl = FALSE}
(temperature <- read.csv('data/temperature.csv'))
```

Wow... that was a lot of output. At least it means the data loaded
properly. Let's check the top (the first 6 lines) of this `data.frame` using the
function `head()`:

```{r, results='show', purl=FALSE}
head(temperature)
```

A `data.frame` is the representation of data in the format of a table where the
columns are vectors that all have the same length. Because each column is a
vector, they all contain the same type of data. We can see this when inspecting
the __str__ucture of a `data.frame` with the function `str()`:

```{r, purl=FALSE}
str(temperature)
```

### Challenge

Based on the output of `str(temperature)`, can you answer the following questions?

* What is the class of the object `temperature`?
* How many rows and how many columns are in this object?
* How many countries have been recorded during these years?

<!---
```{r, echo=FALSE, purl=FALSE}
## Answers
##
## * class: data frame
## * how many rows: 48470,  how many columns: 11
## * how many countries: 8
```
--->

As you can see, many of the columns consist of integers, however, the columns
`City`, `Country` and `country_id` are of a special class called a `factor`. Before we learn
more about the `data.frame` class, let's talk about factors. They are very
useful but not necessarily intuitive, and therefore require some attention.


## Factors

```{r, echo=FALSE, purl=TRUE}
### Factors
```

Factors are used to represent categorical data. Factors can be ordered or
unordered, and understanding them is necessary for statistical analysis and for
plotting.

Factors are stored as integers, and have labels associated with these unique
integers. While factors look (and often behave) like character vectors, they are
actually integers under the hood, and you need to be careful when treating them
like strings.

Once created, factors can only contain a pre-defined set of values, known as
*levels*. By default, R always sorts *levels* in alphabetical order. For
instance, if you have a factor with 2 levels:

```{r, purl=TRUE}
sex <- factor(c("male", "female", "female", "male"))
```

R will assign `1` to the level `"female"` and `2` to the level `"male"` (because
`f` comes before `m`, even though the first element in this vector is
`"male"`). You can check this by using the function `levels()`, and check the
number of levels using `nlevels()`:

```{r, purl=FALSE}
levels(sex)
nlevels(sex)
```

Sometimes, the order of the factors does not matter, other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high") or
it is required by a particular type of analysis. Additionally, specifying the
order of the levels allows for level comparison:

```{r, purl=TRUE, error=TRUE}
food <- factor(c("low", "high", "medium", "high", "low", "medium", "high"))
levels(food)
food <- factor(food, levels=c("low", "medium", "high"))
levels(food)
min(food) ## doesn't work
food <- factor(food, levels=c("low", "medium", "high"), ordered=TRUE)
levels(food)
min(food) ## works!
```

In R's memory, these factors are represented by integers (1, 2, 3), but are more
informative than integers because factors are self describing: `"low"`,
`"medium"`, `"high"`" is more descriptive than `1`, `2`, `3`. Which is low?  You
wouldn't be able to tell just from the integer data. Factors, on the other hand,
have this information built in. It is particularly helpful when there are many
levels (like countries and cities in our example data set).

### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

Converting factors where the levels appear as numbers (such as 
concentration levels) to a numeric vector is a little trickier.  One method is to convert factors to characters and then numbers.  Another method is to use the `levels()` function. Compare:

```{r, purl=TRUE}
f <- factor(c(1, 5, 10, 2))
as.numeric(f)               ## wrong! and there is no warning...
as.numeric(as.character(f)) ## works...
as.numeric(levels(f))[f]    ## The recommended way.
```

Notice that in the `levels()` approach, three important steps occur:

* We obtain all the factor levels using `levels(f)`
* We convert these levels to numeric values using `as.numeric(levels(f))`
* We then access these numeric values using the underlying integers of the
  vector `f` inside the square brackets

### Challenge

The function `table()` tabulates observations and can be used to create bar
plots quickly. For instance, the code below gives you a barplot of the number of
observations.

* In which order are the treatments listed?
* How can you recreate this plot with "control" listed last instead
of first?


```{r wrong-order, results='show', purl=TRUE}
exprmt <- factor(c("treat1", "treat2", "treat1", "treat3", "treat1", "control",
                   "treat1", "treat2", "treat3"))
table(exprmt)
barplot(table(exprmt))
```



```{r correct-order, purl=FALSE}
#Answers

# * The treatments are listed in alphabetical order because they are factors.
# * By redefining the order of the levels
#exprmt <- factor(exprmt, levels=c("treat1", "treat2", "treat3", "control"))
#barplot(table(exprmt))
```
